Variation in gaze following across the life span: A process-level perspective

### 2024-05-14 paper/manuscript.Rmd
improve transition to non-social vector models

affiliation:
  - id            : "1"
    institution   : "Department of Comparative Cultural Psychology, Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany"
  - id            : "2"
    institution   : "Institute of Psychology, Leuphana University Lüneburg, Germany"
  - id            : "3"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "*"
    institution   : "Shared senior authorship" 
abstract: |
  Following eye gaze is fundamental for many social-cognitive abilities, for example, when judging what another agent can or cannot know. While the emergence of gaze following has been thoroughly studied on a group level, we know little about (a) the developmental trajectory beyond infancy and (b) the sources of individual differences. In Study 1, we examined gaze following across the lifespan (*N* = 471 3- to 80-year-olds; children from a mid-sized German city; international remotely tested adults). We found a steep performance improvement during preschool years, in which children became more precise in locating the attentional focus of an agent. Precision levels then stayed comparably stable throughout adulthood with a minor decline toward old age. In Study 2, we formalized the process of gaze following in a computational cognitive model that allowed us to conceptualize individual differences in a psychologically meaningful way (*N* = 60 3- to 5-year-olds, 50 adults). According to our model, participants estimate pupil angles with varying levels of precision based on observing the pupil location within the agent's eyes. In Study 3, we empirically tested how gaze following relates to vector following in non-social settings and perspective-taking abilities (*N* = 102 4- to 5-year-olds). We found that gaze following is associated with both of these abilities but less so with other Theory of Mind tasks. This work illustrates how the combination of reliable measurement instruments and formal theoretical models allows us to explore the in(ter)dependence of core social-cognitive processes in greater detail.
keywords          : "social-cognitive development, theory of mind, gaze following, individual differences, cognitive modeling, lifespan"
wordcount         : "10469"
floatsintext      : yes
figurelist        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes 
draft             : no
classoption       : "man"

@@

**Authors:** Julia Christin Prein^1^, Luke Maurits^1^, Annika Werwach^1,3,4^, Daniel B. M. Haun^1,\*^, Manuel Bohn^1,2,\*^

@@

We collected data online from 3- to 80- year-olds (see Supplements for further details). The child sample consisted of `r sum(lifespan_sample$n_total[lifespan_sample$age_group < 18])` participants and was recruited via an internal database of families in Leipzig, Germany, who volunteered to participate in child development studies. Participants came from ethnically homogeneous, mixed socioeconomic backgrounds with mid to high parental education levels. They lived in an industrialized, urban Central-European context in a mid-size German city (approx. 600,000 inhabitants; median individual monthly net income approx. 1,600€ as of 2021). Most were raised monolingually in a nuclear two-generational family setting. Information on demographics and socioeconomic status was not formally recorded on a participant level.

Adults were recruited via *Prolific* [@palan2018prolific]. *Prolific* is an online participant recruitment tool from the University of Oxford with predominantly European and US-American subjects. Participants consisted of `r sum(lifespan_sample$n_total[lifespan_sample$age_group >= 20])` English-speaking adults who reported to have normal or corrected-to-normal vision. For completing the study, subjects were paid above the fixed minimum wage (~£10.00/hour).

@@

We found very clear support for our gaze model, both in children as well as adults. A strong correlation between the data mean and the gaze model estimate ($\sigma_v$) showed that the data mean is suitable to quantify individual differences: child sample *r*(`r gazemodel_corrs$n[gazemodel_corrs$corr == "gazemodel_param_kids"]`) = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_param_kids"]`,  *p* < .001, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_param_kids"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_param_kids"]`]; adult sample *r*(`r gazemodel_corrs$n[gazemodel_corrs$corr == "gazemodel_param_adults"]`) = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_param_adults"]`,  *p* < .001, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_param_adults"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_param_adults"]`]. The gaze model predicted a U-shaped pattern which we also observed in our data (see Figure \@ref(fig:fig3)C). The model comparison strongly favored our gaze model over the center bias model (child sample $logBF_{10}$ = `r gazemodel_comparison_kids$bf_inf_center`; adult sample $logBF_{10}$ = `r gazemodel_comparison_adults$bf_inf_center`) and the random guessing model (child sample $logBF_{10}$ = `r gazemodel_comparison_kids$bf_inf_guess`; adult sample $logBF_{10}$ = `r gazemodel_comparison_adults$bf_inf_guess`). For the child sample, we went one step further into the analysis. When correlating the observed data across all target positions with the predictions of the three models, we found a high similarity for the gaze model: *r*(`r gazemodel_corrs$n[gazemodel_corrs$corr == "gazemodel_pred_kids"]`) = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_pred_kids"]`,  *p* < .001, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_pred_kids"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_pred_kids"]`], while the correlations with the alternative models were smaller (center bias model: *r*(`r gazemodel_corrs$n[gazemodel_corrs$corr == "centermodel_pred_kids"]`) = `r gazemodel_corrs$r[gazemodel_corrs$corr == "centermodel_pred_kids"]`,  *p* < .001, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "centermodel_pred_kids"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "centermodel_pred_kids"]`]; random guessing model: *r*(`r gazemodel_corrs$n[gazemodel_corrs$corr == "randommodel_pred_kids"]`) = `r gazemodel_corrs$r[gazemodel_corrs$corr == "randommodel_pred_kids"]`,  *p* < .001, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "randommodel_pred_kids"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "randommodel_pred_kids"]`]). The gaze model's prior over potential target locations assumed that participants' clicks would be skewed towards the screen center. For children, we estimated the width of the prior's distribution based on the participants' age. Interestingly, we found that the differential age effects on the center bias prior were minor (intercept = `r gazemodel_prior$mode[gazemodel_prior$type == "intercept"]`, 95% high-density interval (HDI) [`r gazemodel_prior$lci[gazemodel_prior$type == "intercept"]`; `r gazemodel_prior$uci[gazemodel_prior$type == "intercept"]`]; slope = `r gazemodel_prior$mode[gazemodel_prior$type == "slope"]`, 95%HDI [`r gazemodel_prior$lci[gazemodel_prior$type == "slope"]`; `r gazemodel_prior$uci[gazemodel_prior$type == "slope"]`]). The slope of the prior indicated that older children were only marginally less drawn towards the screen center compared to the younger children.

(ref:figlab3) **Gaze model results** (A) Developmental trajectory of th±e estimated model parameter. Grey dots show individual level parameter values. The black line shows the maximum a posteriori (MAP) estimate; blue lines show 1000 draws from it. The large black point with 95% HDI shows the mean of the model parameter for the adult sample. We added minimal horizontal and vertical noise to the adult individual level parameter values to avoid overplotting. (B) Correlation between estimated mode of the model parameter and data mean per individual, faceted by age group. In the child sample, color denotes age in years. Grey regression lines with 95% CI show smooth conditional means based on linear models, with *Pearson*'s correlation coefficient *r*. (C) Pattern recovery. Imprecision in target width for each target bin by age group. Model predictions in blue; data in grey. (D) Correlation between the observed data and the predictions of the three models by target position and age group (across individuals; only for the child sample).

@@

Previous models in the social- and non-social cognitive domain exist that rely on vector estimations (see @madl2015computational for a review of computational models for spatial memory). For example, @brom2012computational defined egocentric self-to-object vectors and allocentric object-to-object vectors where corresponding uncertainties due to memory and motor errors were modelled by vector weights and Gaussian noise. The counterfactual simulation model [@gerstenberg2021counterfactuala, @gerstenberg2021counterfactual, @gerstenberg2012noisy, @gerstenberg2014counterfactual] used vector calculations to model causal reasoning, for example, in estimating billiard ball trajectories and their collisions. Gaussian noise was added to the ball’s movement vector to represent people’s prior expectations, uncertainties, and expertise (the stronger the expectations, the less noise is added to the vector).

In the social domain, there are additional examples of following vectors in space. Recent work on pointing gestures introduced the idea of a touch-line vector [@omadagain2019origin], see @li2023understanding for computational modeling. According to this hypothesis, humans directed their fingers not to construct an "arrow" pointing at the target, but rather as if they aimed to touch it directly. Consequently, the touch line is a vector that connects the eye and the pointed fingertip, and marks the object that the person's fingertip seems closest to touching from her perspective [@omadagain2019origin]. This supports the idea that vector estimation is relevant for social-cognitive abilities.

@@

Interestingly, gaze following and perspective-taking correlated with each other, $\rho$(`r magnet_cormatrix$n[magnet_cormatrix$Correlation == "Tango x Perspective-taking"]`) = `r magnet_cormatrix$r[magnet_cormatrix$Correlation == "Tango x Perspective-taking"]`, *p* = `r magnet_cormatrix$p[magnet_cormatrix$Correlation == "Tango x Perspective-taking"]`, 95%CI [`r magnet_cormatrix$lower[magnet_cormatrix$Correlation == "Tango x Perspective-taking"]`, `r magnet_cormatrix$upper[magnet_cormatrix$Correlation == "Tango x Perspective-taking"]`]. Please note that the TANGO quantifies imprecision in gaze following. Therefore, a negative correlation suggests that imprecision in gaze following corresponds to less perspective-taking. Non-social vector following and perspective-taking did not correlate, $\rho$(`r magnet_cormatrix$n[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`) = `r magnet_cormatrix$r[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`, *p* = `r magnet_cormatrix$p[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`, 95%CI [`r magnet_cormatrix$lower[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`, `r magnet_cormatrix$upper[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`]. However, according to the Williams' test, the two correlations did not differ significantly from each other, *t*(`r magnet_williamsttest_perspective$parameter`) = `r round(magnet_williamsttest_perspective$statistic, 2)`, *p* = `r round(magnet_williamsttest_perspective$p.value, 3)`.

>>>>

affiliation:
  - id            : "1"
    institution   : "Department of Comparative Cultural Psychology, Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany"
  - id            : "2"
    institution   : "Institute of Psychology, Leuphana University Lüneburg, Germany"
  - id            : "3"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "*"
    institution   : "Shared senior authorship" 
abstract: |
  Following eye gaze is fundamental for many social-cognitive abilities, for example, when judging what another agent can or cannot know. While the emergence of gaze following has been thoroughly studied on a group level, we know little about (a) the developmental trajectory beyond infancy and (b) the sources of individual differences. In Study 1, we examined gaze following across the lifespan (*N* = 478 3- to 19-year-olds from a mid-sized German city; and *N* = 240 20- to 80-year-old international, remotely tested adults). We found a steep performance improvement during preschool years, in which children became more precise in locating the attentional focus of an agent. Precision levels then stayed comparably stable throughout adulthood with a minor decline toward old age. In Study 2, we formalized the process of gaze following in a computational cognitive model that allowed us to conceptualize individual differences in a psychologically meaningful way (*N* = 60 3- to 5-year-olds, 50 adults). According to our model, participants estimate pupil angles with varying levels of precision based on observing the pupil location within the agent's eyes. In Study 3, we empirically tested how gaze following relates to vector following in non-social settings and perspective-taking abilities (*N* = 102 4- to 5-year-olds). We found that gaze following is associated with both of these abilities but less so with other Theory of Mind tasks. This work illustrates how the combination of reliable measurement instruments and formal theoretical models allows us to explore the in(ter)dependence of core social-cognitive processes in greater detail.
keywords          : "social-cognitive development, theory of mind, gaze following, individual differences, cognitive modeling, lifespan"
wordcount         : "10469"
floatsintext      : yes
figurelist        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes 
draft             : no
classoption       : "man"

@@

**Authors:** Julia Christin Prein^1,2^, Luke Maurits^1^, Annika Werwach^1,3,4^, Daniel B. M. Haun^1,\*^, Manuel Bohn^1,2,\*^

@@

We collected data online from 3- to 80- year-olds (see Supplements for further details). The child sample consisted of `r sum(lifespan_sample$n_total[lifespan_sample$age_group < 20])` 3- to 19-year-olds and was recruited via an internal database of families in Leipzig, Germany, who volunteered to participate in child development studies. Participants came from ethnically homogeneous, mixed socioeconomic backgrounds with mid to high parental education levels. They lived in an industrialized, urban Central-European context in a mid-size German city (approx. 600,000 inhabitants; median individual monthly net income approx. 1,600€ as of 2021). Most were raised monolingually in a nuclear two-generational family setting. Information on demographics and socioeconomic status was not formally recorded on a participant level.

Adults were recruited via *Prolific* [@palan2018prolific]. *Prolific* is an online participant recruitment tool from the University of Oxford with predominantly European and US-American subjects. Participants consisted of `r sum(lifespan_sample$n_total[lifespan_sample$age_group >= 20])` English-speaking adults between 20 and 80 years of age who reported to have normal or corrected-to-normal vision. For completing the study, subjects were paid above the fixed minimum wage (~£10.00/hour).

@@

We found very clear support for our gaze model, both in children as well as adults. A strong correlation between the data mean and the gaze model estimate ($\sigma_v$) showed that the data mean is suitable to quantify individual differences: child sample *r*(`r gazemodel_corrs$n[gazemodel_corrs$corr == "gazemodel_param_kids"]`) = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_param_kids"]`,  *p* < .001, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_param_kids"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_param_kids"]`]; adult sample *r*(`r gazemodel_corrs$n[gazemodel_corrs$corr == "gazemodel_param_adults"]`) = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_param_adults"]`,  *p* < .001, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_param_adults"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_param_adults"]`]. The gaze model predicted a U-shaped pattern which we also observed in our data (see Figure \@ref(fig:fig3)C). The model comparison strongly favored our gaze model over the center bias model (child sample $logBF_{10}$ = `r gazemodel_comparison_kids$bf_inf_center`; adult sample $logBF_{10}$ = `r gazemodel_comparison_adults$bf_inf_center`) and the random guessing model (child sample $logBF_{10}$ = `r gazemodel_comparison_kids$bf_inf_guess`; adult sample $logBF_{10}$ = `r gazemodel_comparison_adults$bf_inf_guess`). For the child sample, we went one step further into the analysis. When correlating the observed data across all target positions and age groups with the predictions of the three models, we found a high similarity for the gaze model: *r*(`r gazemodel_corrs$n[gazemodel_corrs$corr == "gazemodel_pred_kids"]`) = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_pred_kids"]`,  *p* < .001, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_pred_kids"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_pred_kids"]`], while the correlations with the alternative models were smaller (center bias model: *r*(`r gazemodel_corrs$n[gazemodel_corrs$corr == "centermodel_pred_kids"]`) = `r gazemodel_corrs$r[gazemodel_corrs$corr == "centermodel_pred_kids"]`,  *p* < .001, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "centermodel_pred_kids"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "centermodel_pred_kids"]`]; random guessing model: *r*(`r gazemodel_corrs$n[gazemodel_corrs$corr == "randommodel_pred_kids"]`) = `r gazemodel_corrs$r[gazemodel_corrs$corr == "randommodel_pred_kids"]`,  *p* < .001, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "randommodel_pred_kids"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "randommodel_pred_kids"]`]). The gaze model's prior over potential target locations assumed that participants' clicks would be skewed towards the screen center. For children, we estimated the width of the prior's distribution based on the participants' age. Interestingly, we found that the differential age effects on the center bias prior were minor (intercept = `r gazemodel_prior$mode[gazemodel_prior$type == "intercept"]`, 95% highest-density interval (HDI) [`r gazemodel_prior$lci[gazemodel_prior$type == "intercept"]`; `r gazemodel_prior$uci[gazemodel_prior$type == "intercept"]`]; slope = `r gazemodel_prior$mode[gazemodel_prior$type == "slope"]`, 95%HDI [`r gazemodel_prior$lci[gazemodel_prior$type == "slope"]`; `r gazemodel_prior$uci[gazemodel_prior$type == "slope"]`]). The slope of the prior indicated that older children were only marginally less drawn towards the screen center compared to the younger children.

(ref:figlab3) **Gaze model results** (A) Developmental trajectory of the estimated model parameter. Grey dots show individual level parameter values. The black line shows the maximum a posteriori (MAP) estimate; blue lines show 1000 draws from it. The large black point with 95% HDI shows the mean of the model parameter for the adult sample. We added minimal horizontal and vertical noise to the adult individual level parameter values to avoid overplotting. (B) Correlation between estimated mode of the model parameter and data mean per individual, faceted by age group. In the child sample, color denotes age in years. Grey regression lines with 95% CI show smooth conditional means based on linear models, with *Pearson*'s correlation coefficient *r*. (C) Pattern recovery. Imprecision in target width for each target bin by age group. Model predictions in blue; data in grey. (D) Correlation between the observed data and the predictions of the three models by target position and age group (across individuals; only for the child sample).

@@

Previous models in the social- and non-social cognitive domain exist that also make use of the concept of vector estimation (see @madl2015computational for a review of computational models for spatial memory). For example, @brom2012computational defined egocentric self-to-object vectors and allocentric object-to-object vectors where corresponding uncertainties due to memory and motor errors were modeled by vector weights and Gaussian noise. The counterfactual simulation model [@gerstenberg2021counterfactuala; @gerstenberg2021counterfactual; @gerstenberg2012noisy; @gerstenberg2014counterfactual] used vector calculations to model causal reasoning; for example, in estimating billiard ball trajectories and their collisions. Gaussian noise was added to the ball’s movement vector to represent people’s prior expectations, uncertainties, and expertise (the stronger the expectations, the less noise is added to the vector). Therefore, the way these models capture individual differences resembles the approach taken in the here-proposed gaze model.

The idea of gaze as vector following has recently been introduced to the study of pointing gestures (@omadagain2019origin, see @li2023understanding for computational modeling). According to this hypothesis, when humans use pointing gestures, the index finger does not constitute "arrow" pointing at the target, but is used to touch the object within one's visual field. Consequently, the touch line is a vector that connects the eye and the pointed fingertip, and marks the object that the person's fingertip seems closest to touching from her perspective [@omadagain2019origin]. This supports the idea that vector estimation is relevant for social-cognitive abilities.

@@

Interestingly, gaze following and perspective-taking correlated with each other, $\rho$(`r magnet_cormatrix$n[magnet_cormatrix$Correlation == "Tango x Perspective-taking"]`) = `r magnet_cormatrix$r[magnet_cormatrix$Correlation == "Tango x Perspective-taking"]`, *p* < .01, 95%CI [`r magnet_cormatrix$lower[magnet_cormatrix$Correlation == "Tango x Perspective-taking"]`, `r magnet_cormatrix$upper[magnet_cormatrix$Correlation == "Tango x Perspective-taking"]`]. Please note that the TANGO quantifies imprecision in gaze following. Therefore, a negative correlation suggests that imprecision in gaze following corresponds to less perspective-taking. Non-social vector following and perspective-taking did not correlate, $\rho$(`r magnet_cormatrix$n[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`) = `r magnet_cormatrix$r[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`, *p* = `r magnet_cormatrix$p[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`, 95%CI [`r magnet_cormatrix$lower[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`, `r magnet_cormatrix$upper[magnet_cormatrix$Correlation == "Magnet x Perspective-taking"]`]. However, according to the Williams' test, the two correlations did not differ significantly from each other, *t*(`r magnet_williamsttest_perspective$parameter`) = `r round(magnet_williamsttest_perspective$statistic, 2)`, *p* = `r round(magnet_williamsttest_perspective$p.value, 3)`.
